Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2010-10-22 pkg/paper/Paper/draft.Rnw
Some cleanup in SVN, now three directories:   Analysis   Paper   Slides where the latter two rely on the former.
All figures removed because they are now dynamically generated in .Rnw.

Two target files (draft.pdf, slides.pdf) are retained to be linked from
R-Forge project web page.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@60 b679b2f7-6539-45d0-a605-5ddcb976cde9

\documentclass[nojss]{jss}
\usepackage[latin1]{inputenc}
\usepackage{longtable}
\usepackage{lscape}
\usepackage{multirow}

@@

% Verkaufsstrategien
% 1. Widerlegung des Teuro Effektes
% 2. Keine Veränderung d. EMU sichtbar; nach 2002 kein globaler Trend aller EMU Staaten Richtung Wechsel
% 3. Nutzen zeigen, den die osteuropäischen Staaten haben, dadurch, dass sie dem EURO beitreten müssen  (mit Ausnahme von Slovenien fallender Trend in Mittelwert und Varianz)
% 	 so etwa in Slowenien und der Slovakei ein Bruch bevor diese Staaten dem ERMII/EURO beitreten
% 4. Finanzkrise: schwieriger, ausser Irland keine Änderung > 2008

@@

In econometrics, the logistic distribution is often used in income distributions and growth models. This is due to  its longer tails and  higher peak, which fits these problems somewhat better. The use of the generalized logistic distribution as we will use it in this paper is rather rare. \citet{won} uses a GL distribution in a regression model with autocorrelated errors and assumes that they follow a GL distribution rather than a Student's t-distribution, as to model the fact that these are oftentimes  leptokurtic and severely left or right skewed. A similar GL distribution is also used in \citet{tolikas} who analyses extreme risk and value--at--risk in the German stock market, all though they don't use a shape parameter. Regarding inflation rates, the GL distribution is --- to our best knowledge --- only used in relationship with expected inflation. \citet{batchelor} use a logistic distribution (not its generalization) to model the distribution of mean expected inflation rates. \\
We apply the general framework, as developed in \citet{z07}, to a more specific model, in this case by means of a GL distribution. Prior to that, we would like to give a short justification for the usage of the GL distribution. Regarding the data at hand, it was not possible to use the already existing method developed in \citet{z07}, since almost all inflation rates, with the notable exception of Greece, are not normally distributed and clearly exhibit asymmetric properties. \\
Therefore, a somewhat more flexible distribution had to be used. We needed a distribution exhibiting rather strong kurtosis and the property to be both left and right skewed. To serve this purpose, we use a generalization of the logistic distribution as defined in \citet{johnson}. Its probability density is given by:

@@

To help with the interpretation, we give a short overview over the history of the EMU. Following the Delors Plan with his 3 stages, we have: stage I (1990-1994) as a phase of liberalization, stage II (1995-1998) a phase of convergence and stage III (1999-2002) the transition period, which ended with the introduction of the Euro as legal tender. If the EMU had a significant effect upon inflation rate volatility we would expect to find a break either in the 90ies, reflecting the various waves of integration or after the introduction of the Euro due to the popular argument that the introduction of Euro paper money led to a considerable price increase. 
% http://en.wikipedia.org/wiki/European_Exchange_Rate_Mechanism#Replacement_with_the_euro_and_ERM_II
% http://en.wikipedia.org/wiki/Eurozone

\begin{longtable}{|l|l|l|l|l|l|l|l|l|l|l|l|l|l|l|l|} 
\hline
Country & Segment~1 & Segment~2 & Segment~3 & ERM~II/EMS & EURO \\
\hline
\hline
\multirow{4}{*}{Austria} & 1990(2)--2007(9)   & 2007(10)--2010(3) & -- & 1995(1) & 1999(1) \\ 
\cline{2-4}
 & mean: 0.1635 & mean: 0.1731 & &  &  \\
 & var: 0.05694 & var: 0.16310 & &  &  \\
 & skew: 0.6056 & skew: 0.1691 &   &  &  \\
\hline 
\hline
\multirow{4}{*}{Belgium} & 1991(2)--1999(12)   & 2000(1)--2010(3) & -- & 1979(1) & 1999(1) \\ 
\cline{2-4}
 & mean: 0.1459 & mean: 0.1768 & &  &  \\
 & var: 0.06401 & var: 0.95401 & &  &  \\
 & skew: -0.03708 & skew: 0.50356 &   &  &  \\
\hline 
\hline
\multirow{4}{*}{Czech Republic} & 1995(2)--1998(7)   & 1998(8)--2010(3) & -- & no & no \\ 
\cline{2-4}
 & mean: 0.6969 & mean: 0.1821& &  &  \\
 & var: 0.3363 & var: 0.2156 & &  &  \\
 & skew: 1.139 & skew: 0.990 &   &  &  \\
\hline 
\hline
\multirow{4}{*}{Denmark} & 1990(2)--2000(6)   & 2000(7)--2010(3) & -- & 1999(1) & no \\ 
\cline{2-4}
 & mean: 0.1664 & mean: 0.1676& &  &  \\
 & var: 0.09078 & var: 0.18758 & &  &  \\
 & skew: -0.7425 & skew: 1.0471 &   &  &  \\
\hline 
\hline
\multirow{4}{*}{Estonia} & 1996(2)--1998(3)   & 1998(4)--2010(3) & -- & 2004(6) & 2011(1) \\ 
\cline{2-4}
 & mean: 0.8649 & mean: 0.3328 & &  &  \\
 & var: 0.4196 & var: 0.2062 & &  &  \\
 & skew: 0.4041 & skew: 0.8016 &   &  &  \\
\hline 
\hline
\multirow{4}{*}{Finland} & 1990(2)--2010(3)   & -- & -- & 1996(10) & 1999(1) \\ 
\cline{2-4}
 & mean: 0.1653 &  & &  &  \\
 & var: 0.1321 &  & &  &  \\
 & skew: 0.2798 &  &   &  &  \\
\hline 
\hline
\multirow{4}{*}{France} & 1990(2)--2004(12)   & 2005(1)--2010(3) & -- & 1979(1) & 1999(1) \\ 
\cline{2-4}
 & mean: 0.1588 & mean: 0.1504 & &  &  \\
 & var: 0.05769 & var: 0.13134 & &  &  \\
 & skew: 0.1965 & skew: -0.7942 &   &  &  \\
\hline 
\hline
\multirow{4}{*}{Germany} & 1995(2)--2000(5)   & 2000(6)--2004(12) & 2005(1)--2010(3) & 1979(1) & 1999(1) \\ 
\cline{2-4}
 & mean: 0.08799 & mean: 0.14018 & mean: 0.14183 &  &  \\
 & var: 0.06013 & var: 0.16352 & var: 0.18384 &  &  \\
 & skew: 0.9219 & skew: 0.9920 & skew: -0.6625 &  &  \\
\hline
\newpage 
\hline
\multirow{4}{*}{Greece} & 1995(2)--2010(3)   & -- & -- & 1999(1) & 2001(1) \\ 
\cline{2-4}
 & mean: 0.3227 &  & &  &  \\
 & var: 1.48 &  & &  &  \\
 & skew: 0.4314 &  &   &  &  \\
\hline
\hline
\multirow{4}{*}{Hungary} & 1995(2)--1998(5)   & 1998(6)--2010(3) & -- & no & no \\ 
\cline{2-4}
 & mean: 1.6064 & mean: 0.5068 & &  &  \\
 & var: 1.0243 & var: 0.3161 & &  &  \\
 & skew: 0.8781 & skew: 0.7095 &   &  &  \\
\hline 
\hline
\multirow{4}{*}{Ireland} & 1995(2)--2008(3)   & 2008(4)--2010(3) & -- & 1979(1) & 1999(1) \\ 
\cline{2-4}
 & mean: 0.2546 & mean: -0.1313 & &  &  \\
 & var: 0.2045 & var: 0.1836 & &  &  \\
 & skew: -0.6958 & skew: -0.9947 &   &  &  \\
\hline 
\hline
\multirow{4}{*}{Italy} & 1990(2)--1996(5)   & 1996(6)--2000(12) & 2001(1)--2010(3) & 1979(1) & 1999(1) \\ 
\cline{2-4}
 & mean: 0.4135 & mean: 0.1676 & mean: 0.1819&  &  \\
 & var: 0.04129 & var: 0.01997 & var: 0.32117 &  &  \\
 & skew: 0.9627 & skew: 0.7261 & skew: -0.2605 &  &  \\
\hline 
\hline
\multirow{4}{*}{Luxembourg} & 1995(2)--1998(12)   & 1999(1)--2010(3) & -- & 1979(1) & 1999(1) \\ 
\cline{2-4}
 & mean: 0.08761 & mean: 0.22425 & &  &  \\
 & var: 0.01340 & var: 0.53108 & &  &  \\
 & skew: 0.2606 & skew: -0.4836 &   &  &  \\
\hline 
\hline
\multirow{4}{*}{Netherlands} & 1990(2)--2010(3)   & -- & -- & 1979(1) & 1999(1) \\ 
\cline{2-4}
 & mean: 0.1854 &  & &  &  \\
 & var: 0.293 &  & &  &  \\
 & skew: 0.5984 &  &   &  &  \\
\hline 
\hline
\multirow{4}{*}{Poland} & 1996(2)--2001(5)   & 2001(6)--2010(3) & -- & no & no \\ 
\cline{2-4}
 & mean: 0.8548 & mean: 0.2024 & &  &  \\
 & var: 0.4212 & var: 0.1232 & &  &  \\
 & skew: 0.6675 & skew: -0.3148 &   &  &  \\
\hline 
\hline
\multirow{4}{*}{Portugal} & 1990(2)--1992(7)   & 1992(8)--2004(3) & 2004(4)--2010(3) & 1992(4) & 1999(1) \\ 
\cline{2-4}
 & mean: 0.8519 & mean: 0.2700 & mean: 0.1605&  &  \\
 & var: 0.1718 & var: 0.1052 & var: 0.2559 &  &  \\
 & skew: 1.1394 & skew: 0.8653 & skew: 0.5690 &  &  \\
\hline 
\hline
\multirow{4}{*}{Slovakia} & 1995(2)--1997(4)   & 1997(5)--2004(2) & 2004(3)--2010(3) & 2005(11) & 2009(1) \\ 
\cline{2-4}
 & mean: 0.4799 & mean: 0.5872 & mean: 0.1865 &  &  \\
 & var: 0.05768 & var: 0.44150 & var: 0.08903 &  &  \\
 & skew: 1.139 & skew: 1.140 & skew: 1.139 &  &  \\
\hline 
\hline
\multirow{4}{*}{Slovenia} & 1996(2)--2003(7)   & 2003(8)--2010(3) & -- & 2004(6) & 2007(1) \\ 
\cline{2-4}
 & mean: 0.6309 & mean: 0.2436 & &  &  \\
 & var: 0.2108 & var: 0.3440 & &  &  \\
 & skew: 0.5883 & skew: 0.1426 &   &  &  \\
\hline 
\newpage
\hline
\multirow{4}{*}{Spain} & 1992(2)--1996(5)   & 1996(6)--2000(12) & 2001(1)--2010(3) & 1989(6) & 1999(1) \\ 
\cline{2-4}
 & mean: 0.3724 & mean: 0.2001 & mean: 0.2225 &  &  \\
 & var: 0.06960 & var: 0.03988 & var: 0.34215 &  &  \\
 & skew: 1.13909 & skew: 0.01874 & skew: -0.36200 &  &  \\
\hline 
\hline
\multirow{4}{*}{Sweden} & 1990(2)--1993(1)   & 1993(2)--2010(3) & -- & no & no \\ 
\cline{2-4}
 & mean: 0.4748 & mean: 0.1552 & &  &  \\
 & var: 0.5715 & var: 0.1848 & &  &  \\
 & skew: 1.1393 & skew: 0.5344 &   &  &  \\
\hline 
\hline
\multirow{4}{*}{United Kingdom} & 1990(2)--1992(4)   & 1992(5)--2010(3) & -- & 1990(10)\footnote{left after the Black Wednesday 1992} & no \\ 
\cline{2-4}
 & mean: 0.5703 & mean: 0.1615 & &  &  \\
 & var: 0.3869 & var: 0.1490 & &  &  \\
 & skew: 1.139 & skew: -1.265 &   &  &  \\
\hline 
\end{longtable}

% \endfoot beendet ganz

@@

\newpage

On the basis of the results, we can trace out 6 groups of countries with rather distinct behavior. Group 1 consists of Italy and Spain (both very similar) and possibly Belgium, Luxembourg and Portugal. Group 2 is an Eastern European group comprising Poland, the Czech Republic, Slovenia, Slovakia, Estonia and Hungary. Group 3 is a Non-ERM~II member group including Denmark, Sweden and the United Kingdom. The Netherlands, Greece and Finland build group 4 which exhibits no changes. Group 5 is composed of France, Germany, and Austria. Ireland stands on its own and is the only country which changed its inflation dynamics as an after march of the financial crises of 2008. \\
What we can observe for almost all the EURO countries is an increase in variance and mean, with the exception of Ireland. Since inflation volatility increased in all countries, we find no proof of a common claim that inflation uncertainty may increase in countries that have a smaller influence on ECB policy. \\
Lets consider a few countries in broader detail to get an idea about what happened. In Austria we see evidence of a change in the inflation rate, when the old regime ended in September 2007. The new regime is much more volatile, with an almost threefold increase in variance.

% sizes !!!

\begin{figure}[ht!p]
  \centering
    \includegraphics{results-012.pdf}
  \caption{Series and supLM test for Austria}
  \label{aut}
\end{figure}

The reason for this is a strong increase in the yearly Austrian inflation rate following a heavy increase in oil price which was further amplified by a rise of mineral taxes. However, we fail to find any evidence for an incrase in mean or volatility of the Austrian rate following the introduction of the Euro paper money. \\
Lets consider Slovenia as the first Eastern European country that introduced the Euro. In August 2003 the new regime started with a much lower mean value (only a third now of its previous value) and interestingly enough, a higher volatility.

\begin{figure}[ht!p]
  \centering
    \includegraphics{results-131.pdf}
  \caption{Series and supLM test for Slovenia}
  \label{aut}
\end{figure}

%
%The LWZ criteria favors one break. 
%
%
%\begin{figure}[ht!p]
%  \centering
%    \includegraphics{results-132.pdf}
%  \caption{LWZ and Negative Log-Likelihood for Break Point Selection}
%  \label{aut}
%\end{figure}

@@

%As a last example we focus on Spain. \\

%Other possible explanations: 
%
%- rise in value added tax (VAT; as in Germany in 2007)
%- change in central bank policy (for example Spain changed to inflation targeting in 1995; Uk, Swe and Finland in 1992)
%- role of budget deficits

% Quelle für diese Behaupunten: http://www.oenb.at/isaweb/dyna2.do?go=initIndikatoren&definitionLaden=false&hierarchieIdSelected=108

>>>>

\documentclass[nojss]{jss}
\usepackage[latin1]{inputenc}
\usepackage{longtable}
\usepackage{lscape}
\usepackage{multirow}
%% need no \usepackage{Sweave}

\setkeys{Gin}{width=\textwidth}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide, keep.source=TRUE}

<<preliminaries>>=
## packages
library("glogis")
library("fxregime")

## additional functions
moments <- function(object, ...) UseMethod("moments")
moments.glogisfit <- function(object, ...) object$moments
moments.breakpoints.glogisfit <- function(object, breaks = NULL, ...)
  t(sapply(refit(object, breaks = breaks), "[[", "moments"))

## For the creation of the tables
## <<tabelle, eval=FALSE>>=
if(!file.exists(file.path("C:/Dissertation/Paper-1-Z", "Analysis", "Austria.rda"))) {
setwd(file.path("..", "Analysis"))
Sweave("analysis.Rnw")
setwd(file.path("..", "Slides"))
}

## query result file names
fnam <- Sys.glob(file.path("..", "Analysis", "*.rda"))
fnam2 <- grep("2", fnam, fixed = TRUE)
if(length(fnam2) > 0) fnam <- fnam[-fnam2]

## set up named results list
results <- vector(length = length(fnam), mode = "list")
names(results) <- sapply(strsplit(sapply(strsplit(fnam, "/", fixed = TRUE), tail, 1), ".", fixed = TRUE), head, 1)

## erm <- yearmon(c(
##   Austria = 1995,
##   ...,
##   France = 1995 + 1/12,
##   ...))

## create erm (Exchange rate mechanism bzw. exchange rate mechanism 2
## ERM~II, ERM~I (part of european monetary system expired in 1.1.1999 gave way to ERM II
## if never entered ""right?
## http://web.archive.org/web/20080101022356/http://lexikon.meyers.de/meyers/Europ%C3%A4isches_W%C3%A4hrungssystem

erm <- yearmon(c(
	Austria = 1995,
	Belgium = 1979 + 2/12,
	CzechRepubic = NA, 
	Denmark = 1979 + 2/12,
	Estonia = NA,
	Finland = 1996 + 9/12,
	France = 1979 + 2/12,
	Germany = 1979 + 2/12,
	Greece = 1998 + 2/12,
	Hungary = NA,
	Ireland = 1979 + 2/12,
	Italy = 1979 + 2/12,
	Luxembourg = 1979 + 2/12,
	Netherlands = 1979 + 2/12,
	Poland = NA,
	Portugal = 1992 + 3/12,
	Slovakia = NA,
	Slovenia = NA,
	Spain = 1986 + 5/12,
	Sweden = NA ,
	UnitedKingdom = NA
))

erm2 <- yearmon(c(
	Austria = NA,
	Belgium = NA,
	CzechRepubic = NA, 
	Denmark = 1999,
	Estonia = 2004 + 5/12,
	Finland = NA,
	France = NA,
	Germany = NA,
	Greece = 1999,
	Hungary = NA,
	Ireland = NA,
	Italy = NA,
	Luxembourg = NA,
	Netherlands = NA,
	Poland = NA,
	Portugal = NA,
	Slovakia = 2005 + 10/12,
	Slovenia = 2004 + 5/12,
	Spain = NA,
	Sweden = NA ,
	UnitedKingdom = NA
))

eur <- yearmon(c(
	Austria = 1999,
	Belgium = 1999,
	CzechRepubic = NA, 
	Denmark = NA,
	Estonia = 2011,
	Finland = 1999,
	France = 1999,
	Germany = 1999,
	Greece = 2001,
	Hungary = NA,
	Ireland = 1999,
	Italy = 1999,
	Luxembourg = 1999,
	Netherlands = 1999,
	Poland = NA,
	Portugal = 1999,
	Slovakia = 2009,
	Slovenia = 2007,
	Spain = 1999,
	Sweden = NA ,
	UnitedKingdom = NA
))

## sequentially load data
for(i in seq_along(fnam)) {
  load(file.path("..", "Analysis", fnam[i]))
  results[[i]] <- list(
    data = x,
    model0 = x_gf,
    efp = x_efp,
    nbreaks = x_nbreaks,
    breakpoints = x_bp,
    model = x_rf,
    erm = erm[i],
    erm2 = erm2[i],
    eur = eur[i])
}

## compute moments
mom <- lapply(results, function(obj) t(sapply(obj$model, moments)))

## LaTeX formatting
format_moments <- function(i, digits = 3) {
  rval <- format(round(mom[[i]], digits = digits), nsmall = digits)
  rval <- cbind("", rownames(rval), matrix(paste("$", rval, "$", sep = ""), ncol = 3), "", "", "")
  rval[1,1] <- if(is.character(i)) i else names(mom)[i]
  rval[1,6] <- if(is.na(results[[i]]$erm)) "--" else format(results[[i]]$erm)
  rval[1,7] <- if(is.na(results[[i]]$erm2)) "--" else format(results[[i]]$erm2)
  rval[1,8] <- if(is.na(results[[i]]$eur)) "--" else format(results[[i]]$eur)
  rval <- apply(rval, 1, paste, collapse = " & ")
  rval <- paste(rval, "\\\\")
  rval[length(rval)] <- paste(rval[length(rval)], "\\hline")
  return(rval)
}

## data
data("HICP", package = "glogis")
HICP <- HICP[, colnames(HICP) != "EU"]
@
%
%% analysis "macro" (assumes existence of "i", "HICP", and "add_breaks")
<<analysis, eval=FALSE>>=
## compute returns
x <- 100 * diff(log(na.omit(HICP[,i])))

## fit full-sample model
x_gf <- glogisfit(x)

## assess stability of full-sample model
x_efp <- gefp(x_gf, fit = NULL)

## estimate breakpoints (with minimal segment size of 2 years)
x_bp <- breakpoints(x_gf, h = 24)

## (minimal) number of breaks
x_nbreaks <- if(sctest(x_efp, functional = supLM(0.1))$p.value > 0.05) {
  0
} else {
  max(1, length(breakdates(x_bp))) + add_breaks[i]
}

## refit segmented model
x_rf <- refit(x_bp, breaks = x_nbreaks)

## save everything
save(x, x_gf, x_efp, x_bp, x_nbreaks, x_rf, file = paste(colnames(HICP)[i], ".rda", sep = ""))

<<bla, eval=TRUE>>=
setwd("C:/Dissertation/Paper-1-Z/R-Forge2/pkg/paper/Analysis")
load("Belgium.Rda")
xbel <- x
xbel_bp <- x_bp
xbel_efp <- x_efp
xbel_gf <- x_gf
xbelnbreaks <- x_nbreaks
xbel_rf <- x_rf
load("Luxembourg.Rda")

@

<<plotversuch, eval=FALSE>>=
par(mfrow = c(2, 1))
plot(x, main = "Series with Fitted Mean", ylab = "100 * diff(log(HICP))", xlab = "Time")
lines(x_bp, breaks = x_nbreaks)
lines(fitted(x_bp, breaks = x_nbreaks, type = "mean"), col = 4)
plot(xbel, main = "Series with Fitted Mean", ylab = "100 * diff(log(HICP))", xlab = "Time")
lines(xbel_bp, breaks = xbelnbreaks)
lines(fitted(xbel_bp, breaks = xbelnbreaks, type = "mean"), col = 4)
@

% plot "macros" (assuming existence "x", "x_gf", ...)
<<plot1, eval=FALSE>>=
par(mfrow = c(1, 2))
plot(x, main = "Series with Fitted Mean", ylab = "100 * diff(log(HICP))", xlab = "Time")
lines(x_bp, breaks = x_nbreaks)
lines(fitted(x_bp, breaks = x_nbreaks, type = "mean"), col = 4)
plot(x_efp, functional = supLM(0.1), main = "supLM test")
lines(x_bp, breaks = x_nbreaks)
@

<<plot1d, eval=FALSE>>=
plot(x, main = "Slovenia", ylab = "100 * diff(log(HICP))", xlab = "Time")
@

<<plot1b, eval=FALSE>>=
plot(x_efp, functional = supLM(0.1), main = "supLM test")
lines(x_bp, breaks = x_nbreaks)
@

<<plot2, eval=FALSE>>=
par(mfrow = c(1, 1))
plot(x_bp)
@

<<plot3, eval=FALSE>>=
par(mfrow = c(1, x_nbreaks + 1))
for(i in 1:(x_nbreaks+1)) plot(x_rf[[i]], xlab = paste(names(x_rf)[i], "\n", "Goodness-of-fit p-value: ",
  format.pval(summary(x_rf[[i]])$chisq.test$p.value, digits = 4), sep = ""))
@

<<plot4, eval=FALSE>>=
par(mfrow = c(1, 1))
plot(x_efp, aggregate = FALSE, ylim=c(-2.1,2.1))
@

@@

@@

In econometrics, the logistic distribution is often used in income distributions and growth models. This is due to  its longer tails, which fits these problems somewhat better. The use of the generalized logistic distribution as we will use it in this paper is rather rare. \citet{won} uses a GL distribution in a regression model with autocorrelated errors and assumes that they follow a GL distribution rather than a Student's t-distribution, as to model the fact that these are oftentimes  leptokurtic and severely left or right skewed. A similar GL distribution is also used in \citet{tolikas} who analyses extreme risk and value--at--risk in the German stock market, all though they don't use a shape parameter. Regarding inflation rates, the GL distribution is --- to our best knowledge --- only used in relationship with expected inflation. \citet{batchelor} use a logistic distribution (not its generalization) to model the distribution of mean expected inflation rates. \\
We apply the general framework, as developed in \citet{z07}, to a more specific model, in this case by means of a GL distribution. Prior to that, we would like to give a short justification for the usage of the GL distribution. Regarding the data at hand, it was not possible to use the already existing method developed in \citet{z07}, since almost all inflation rates, with the notable exception of Greece, are not normally distributed and clearly exhibit asymmetric properties. \\
Therefore, a somewhat more flexible distribution had to be used. We needed a distribution exhibiting rather fatter tails and the property to be both left and right skewed. To serve this purpose, we use a generalization of the logistic distribution as defined in \citet{johnson}. Its probability density is given by:

@@

To help with the interpretation, we give a short overview over the history of the EMU. Following the Delors Plan with his 3 stages, we have: stage I (1990-1994) as a phase of liberalization, stage II (1995-1998) a phase of convergence and stage III (1999-2002) the transition period, which ended with the introduction of the Euro as legal tender and started with the Exchange Rate Mechanism~II (ERM~II) for Non--members and the irrevocable fixing of exchange rates of Member States. If the EMU had a significant effect upon inflation rate volatility we would expect to find a break either in the 90ies, reflecting the various waves of integration or after the introduction of the Euro due to the popular argument that the introduction of Euro paper money led to a considerable price increase. 
% http://en.wikipedia.org/wiki/European_Exchange_Rate_Mechanism#Replacement_with_the_euro_and_ERM_II
% http://en.wikipedia.org/wiki/Eurozone

% automatic longtable provided by R

\small
\begin{longtable}{|l|l|l|l|l|l|l|l|}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & EURO \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(seq(1,21)), format_moments)))
@
\end{longtable}

@@

\newpage
\subsection{Belgium and Luxembourg}

% to show both series with breakpoints and means
% does not work
%\begin{center}
%<<belluxvers, fig=TRUE, height=7, width=8>>=
%<<plotversuch>>
%@
%\end{center}

On the basis of these results, we are able to trace out different groups. The first of this group consits of Belgium and Luxembourg. This may come as no surprise since both countries formed a monetary union prior even to the start of the European Rate Mechanism (ERM). While both countries enjoyed very stable inflation rates with low values of both mean and variance, this changed around the timing of the irrevocable fixing of the exchange rates. A causal connection to this events seems likely.

% heavy increase in variance both countries exactly around EURO introduction 
% interesting thing: both were allready in a currency union (the belgish--luxembourg union)
\small
\begin{longtable}{|l|l|l|l|l|l|l|l|}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & EURO \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(2, 13), format_moments)))
@
\end{longtable}

% could be looking much better
\begin{center}
<<bellux, fig=TRUE, height=8, width=8>>=
print(xyplot(cbind(100*diff(log(na.omit(HICP[, "Belgium"]))), 100*diff(log(na.omit(HICP[, "Luxembourg"])))), xlab="Time", ylab="", main="", screens=list("Belgium", "Luxembourg")), ylim=c(-2,2))
@
\end{center}

\newpage
\subsection{The Eastern Countries}
\normalsize

The second of these groups are the Eastern European Countries, the Czech Republic, Estonia, Hungary, Poland, Slovakia and Slovenia. In all of these countries, the mean and the variance of their inflation rates declined in the later part of the 90ies. Most of them experienced a break in 1997/1998, with Poland, Slovakia and Slovenia somewhat later.

\small
\begin{longtable}{|l|l|l|l|l|l|l|l|}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & EURO \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(5,10,15,17,18), format_moments)))
@
\end{longtable}

\begin{center}
<<eastern, fig=TRUE, height=7, width=8>>=
print(xyplot(cbind(100*diff(log(na.omit(HICP[, "CzechRepublic"]))), 100*diff(log(na.omit(HICP[, "Estonia"]))), 100*diff(log(na.omit(HICP[, "Hungary"]))), 100*diff(log(na.omit(HICP[, "Poland"]))), 100*diff(log(na.omit(HICP[, "Slovakia"]))), 100*diff(log(na.omit(HICP[, "Slovenia"])))), xlab="Time", ylab="", main="", screens=list("Czech Republic", "Estonia", "Hungary", "Poland", "Slovakia", "Slovenia")))
@
\end{center}

%, ylim=c(-2.7,2.7)

\newpage
\subsection{Italy and Spain}
\normalsize

Another combination of countries with -- this time -- identical results are the two biggest southern European countries Italy and Spain. Both of them experienced high rates of inflation in the early 90ies. Afterwards, in the phase leading up to the monetary union, mean and variance were significantly reduced. This trend changed drastically soon after the fixing of the exchange rates leading to higher mean and much higher variances then ever before since the beginning of the HICP series.

\small
\begin{longtable}{|l|l|l|l|l|l|l|l|}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & EURO \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(12, 19), format_moments)))
@
\end{longtable}

% could be looking much better
\begin{center}
<<itaspa, fig=TRUE, height=7, width=8>>=
print(xyplot(cbind(100*diff(log(na.omit(HICP[, "Italy"]))), 100*diff(log(na.omit(HICP[, "Spain"])))), xlab="Time", ylab="", main="", screens=list("Italy", "Spain"), ylim=c(-2,2)))
@
\end{center}

\newpage
\normalsize
\subsection{Ireland}

Ireland can be considered a rather special case. It is the only economy where the effects of the ongoing financial crisis of 2008 are visible. We find a structural break in 2008 following the strong contraction of the Irish housing bubble and a beginning deflation.

\small
\begin{longtable}{|l|l|l|l|l|l|l|l|}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & EURO \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(11), format_moments)))
@
\end{longtable}

\begin{center}
<<ireland, fig=TRUE, height=7, width=8>>=
plot(100*diff(log(na.omit(HICP[, "Ireland"]))), main="Ireland", ylab="", xlab ="Time", ylim=c(-2,2))
@
\end{center}

\newpage
\normalsize
\subsection{No change countries}

Another interesting group consists of 3 economically very different countries. In all of them we find no evidence of a structural change. Whereas the Netherlands and Finland are both on the lower side of the inflation margin, Greece exhibits some very strange behaviour. It has by far the highest variance of all series and seems to be completely different from all the others, just like some randomly generated series.

\small
\begin{longtable}{|l|l|l|l|l|l|l|l|}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & EURO \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(9,14,6), format_moments)))
@
\end{longtable}

\begin{center}
<<nochange, fig=TRUE, height=7, width=8>>=
print(xyplot(cbind(100*diff(log(na.omit(HICP[, "Finland"]))), 100*diff(log(na.omit(HICP[, "Greece"]))), 100*diff(log(na.omit(HICP[, "Netherlands"])))), xlab="Time", ylab="", main="", screens=list("Finland", "Greece", "Netherlands"), ylim=c(-2.7,2.7)))
@
\end{center}

\newpage
\normalsize
\subsection{The Non--participating Countries}

The last group with clear patterns can be seen with the two countries that decided not to be part of any European Monetary System\footnote{If we ignore the short participating time of the United Kingdom in the EMS}, the United Kingdom and Sweden. Both have a clear break in the early 90ies, and both of these breaks can be traced back to crisis, the currency crisis of the United Kingdom with the "`Black Wednesday"' in September 1992 and the run on the Swedish currency in 1992.

\small
\begin{longtable}{|l|l|l|l|l|l|l|l|}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & EURO \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(20, 21), format_moments)))
@
\end{longtable}

% could be looking much better
\begin{center}
<<sweuk, fig=TRUE, height=7, width=8>>=
print(xyplot(cbind(100*diff(log(na.omit(HICP[, "Sweden"]))), 100*diff(log(na.omit(HICP[, "UnitedKingdom"])))), xlab="Time", ylab="", main="", screens=list("Sweden", "United Kingdom")), ylim=c(-2,2))
@
\end{center}

\newpage
\normalsize
\subsection{The Other Countries}

No clear message is available for: Denmark, France, Germany and Portugal.

\subsubsection{Austria}

Lets consider a few countries in broader detail to get an idea about what happened. In Austria we see evidence of a change in the inflation rate, when the old regime ended in September 2007. The new regime is much more volatile, with an almost threefold increase in variance.

% sizes !!!
<<>>=
i <- 1
nam <- colnames(HICP)[i]
fnam <- paste(nam, ".rda", sep = "")
@
%\section*{\Sexpr{nam}}
<<>>=
if(file.exists(fnam)) {
load(fnam)
} else {
<<analysis>>
}
@
\begin{center}
<<autplot, fig=TRUE, height=4, width=8>>=
<<plot1>>
@
\end{center}

The reason for this is a strong increase in the yearly Austrian inflation rate following a heavy increase in oil price which was further amplified by a rise of mineral taxes. However, we fail to find any evidence for an incrase in mean or volatility of the Austrian rate following the introduction of the Euro paper money. \\

\subsubsection{Slovenia}

Lets consider Slovenia as the first Eastern European country that introduced the Euro. In August 2003 the new regime started with a much lower mean value (only a third now of its previous value) and interestingly enough, a higher volatility.

<<>>=
i <- 18
nam <- colnames(HICP)[i]
fnam <- paste(nam, ".rda", sep = "")
@
%\section*{\Sexpr{nam}}
<<>>=
if(file.exists(fnam)) {
load(fnam)
} else {
<<analysis>>
}
@
\begin{center}
<<slovplot, fig=TRUE, height=4, width=8>>=
<<plot1>>
@
\end{center}

@@

% Quelle für diese Behaupunten: http://www.oenb.at/isaweb/dyna2.do?go=initIndikatoren&definitionLaden=false&hierarchieIdSelected=108
\newpage
\subsection{Alternative Explanations}
Another possible explanation pattern could be the growth in the  money supply. A drawback is the current lack of data\footnote{hopefully to be solved}. But for some countries we can find some plausible findings. Regarding the case of Denmark, the M1 money supply started contracting in June 2000 (on a monthly base). In Estonia we find the highpoint in M3 expansion ending with March 1998, with much lower M3 growth rates afterwards. The same happended to Hungary in April and May 1998. In the Polnish case we find a severe contraction of the money supply in May 2001. Similar findings hold true for Germany as well. \\
In Spain we see a stronger increase in M1 in May and June 1996 (nothing for the other break, data scarceity). In Portugal we find a modest increase in year on year M1 money supply growth starting in July 1992. We find nothing peculiar for Slovakia and Hungary. In Estonia the M1 money supply starts to contract severely on a yearly base following March 1998. In Slovenia we find a contraction of the Money Supply in July 2003, with the whole year 2003 witnessing a sharp decrease in the year on year growth. Poland starts contracting the year on year M1 money supply after June 2000 with very low growth rates till January 2002.
